Semantic integration of physiology phenotypes with an application to the Cellular Phenotype Ontology

### 2012-03-18 paper/drinking-bioinformatics.tex

The large volume and diversity of phenotypes within different species
and across multiple scales and levels of granularity necessitates the
application of flexible strategies for managing and integrating data
so that it becomes amenable to automated comparative analyses. To
integrate biomedical data across heterogeneous information systems,
biomedical ontologies are being developed \cite{Smith2007}. An
ontology is an explicit specification of a conceptualization of a
domain and can be used to make the meaning of terms in a vocabulary
explicit \cite{Gruber1995, Guarino1998}. They play a crucial role in
the annotation of biomedical data and the integration of model
organism databases \cite{go2010, Bada2004, goble}.

@@

In the domain of phenotypes, multiple ontologies have been
developed. In particular, ontologies to characterize mammalian
\cite{Smith2004}, human \cite{Robinson2008}, yeast \cite{ypo} and worm
\cite{wpo} phenotypes are now available, while several more phenotype
ontologies are under development. To benefit from automated reasoning,
integrate phenotypes across species and reuse the content of anatomy
and process ontologies, we have defined process classes using the
framework of the Phenotypic Attribute and Trait (PATO) ontology
\cite{Gkoutos2005}. According to the PATO framework, a phenotype can
be decomposed, using an Entity-Quality model, into an affected entity
and a quality that characterizes {\em how} the entity is affected
\cite{Gkoutos2005}. Such decompositions have been created for several
widely used phenotype ontologies \cite{Mungall2010, Gkoutos2009b,
  obml2011h1}, and are being applied together with methods for reusing
knowledge contained in anatomy ontologies \cite{Mungall2010,
  Hoehndorf2010phene}.

@@

To illustrate these challenges, consider physiological processes of
the heart. One of the heart's functions is {\em Heart beating}, i.e.,
a capability that is realized through processes of the type {\em Heart
  beating}. {\em Blood} is a participant of {\em Heart beating}
processes.  An abnormal phenotype of an organism could be that the
{\em rate of heart beating} is increased. The intended meaning of such
a description is that the number of {\em Heart beating} processes in a
given time interval is higher than normal. Another important attribute
of heart physiology is the fluid flow rate through the heart. For an
abnormal phenotype such as {\em increased rate of fluid flow through
  the heart}, the intended meaning could either be that the amount of
fluid that is moved through the heart within a single heart beating
process is increased or that the amount of fluid that is moved through
the heart within a period of time interval is increased.

Based on these examples, we can make several observations about
process attributes. First, for a process like {\em Heart beating}, we
can distinguish between single occurrences and processes in which {\em
  Heart beating} occurs multiple times. Only the latter kind of
process may have a {\em rate of heart beating}, while {\em Heart
  beating} processes do not have such an attribute. Second, we can
distinguish between abnormal fluid flow rates in {\em Heart beating}
processes and rate of fluid flow through the heart within a given
duration. Both may have entirely different underlying causes and it is
therefore important to distinguish between them. Finally, we may be
able to infer some phenotypes from others, thereby limiting the number
of phenotypes that must be experimentally observed. For example, when
the fluid flow rate in single heart beating processes is increased and
the {\em rate of heart beating} is increased within a process $P$,
then the rate of fluid flow through the heart will be increased for
$P$.

Here, we present the foundations for an ontology of process
phenotypes. We present the outline of a theory in which several kinds
of process attributes can be distinguished so that normal and abnormal
physiology of biological systems can be formally characterized. We
apply this theory to cellular processes and create the Cell Phenotype
Ontology (CPO). The CPO is linked to reference ontologies for
qualities, biological processes, functions and cell components, and
its prime application is the unification of phenotypes on the cellular
level across different species as well as for annotation of cellular
phenotypes in domains in which no such ontology exists.

@@

PATO allows for the description of affected entities by combining
various ontologies that describe the entities that have been affected,
such as the various anatomical ontologies, GO \cite {Ashburner2000b},
the Cell Type Ontology \cite {Bard2005}, SO \cite {Eilbeck2005} etc
with the various qualities it provides for defining how these entities
were affected.  PATO can be used for annotation either directly in a
so called post-composed (post-coordinated) manner or for providing
formal (logical) definitions (equivalence axioms) to ontologies
containing a set of precomposed (pre-coordinated) phenotype terms. For
instance, to describe the decrease in the length of the sexual cycle
of female animals, we can combine the PATO term \textit{decreased
  duration} ({\tt PATO:0000499}) with the Gene Ontology term
\textit{estrous cycle} ({\tt GO:0042698}), whilst if such a term
existed in a pre-composed ontology (for example the {\tt MP:0009007}
term from the Mammalian Phenotype) it could be used to provide an
equivalence statement between that class and the above PATO-based
description.
\end{methods}

@@

Amongst abnormal physiology, we distinguish between abnormalities in a
{\em single occurrence} of a cell component's functioning and an
abnormal {\em pattern of multiple occurrences} of a cell component's
functioning. For example, abnormalities in cell division resulting in
{\em Aneuploidy} refer to abnormalities of {\em cell division}
processes, while an {\em increased rate of cell division} refers to an
abnormality in the pattern of occurrence of multiple cell division
processes. In the CPO we follow the GO and represent abnormalities in
the pattern of occurrence of $X$ as abnormalities of {\em regulation
  of $X$} processes. In particular, an {\em increased rate of cell
  division} is not an attribute of {\em cell division} processes, but
rather an attribute of the {\em regulation of cell division}.

@@

A further domain that will greatly benefit from the CPO is {\em
  systems microscopy}, which aims to understand complex and dynamic
cellular systems by combining automated fluorescence microscopy, cell
microarray platforms, quantitative image analysis and data mining
\cite{Lock2010}.  If we consider some of the studies, which have been
published in this field in the last few years \cite{Neumann2010,
  Schmitz2010, Fuchs2010}, the need for CPO becomes evident.  In the
three studies, live-cell imaging assays and RNAi knockdown were used
to generate phenotypic profiles that quantify the cellular response to
a given siRNA thus allowing identification of hundreds of genes
involved in diverse biological functions including cell division,
migration and survival.  In each study, several phenotypes were
detected and described by the authors without the use of ontologies or
controlled vocabularies, making the integration between datasets
extremely difficult. For example, it is evident that cell division
phenotypes were observed in all three datasets and referred to by
terms such as {\em Mitotic delay/arrest}, {\em Prolonged mitotic
  exit}, {\em Methaphase delay} and {\em Methaphase cells}). Without a
controlled vocabulary of cellular phenotypes, the overlap between such
phenotype descriptions is unclear.

@@

\bibliographystyle{natbib}
\bibliography{/home/leechuck/Documents/papers/bibtex/lc}

>>>>

The large volume and diversity of phenotypes within different species
and across multiple scales and levels of granularity necessitates the
application of flexible strategies for managing and integrating data
so that it becomes amenable to automated comparative analyses. To
integrate biomedical data across heterogeneous information systems,
biomedical ontologies are being developed \cite{Smith2007short}. An
ontology is an explicit specification of a conceptualization of a
domain and can be used to make the meaning of terms in a vocabulary
explicit \cite{Gruber1995, Guarino1998}. They play a crucial role in
the annotation of biomedical data and the integration of model
organism databases \cite{go2010, Bada2004, goble}.

@@

In the domain of phenotypes, multiple ontologies have been
developed. In particular, ontologies to characterize mammalian
\cite{Smith2004}, human \cite{Robinson2008}, yeast \cite{aposhort} and
worm \cite{wpo} phenotypes are now available, while several more
phenotype ontologies are under development. To benefit from automated
reasoning, integrate phenotypes across species and reuse the content
of anatomy and process ontologies, we have defined process classes
using the framework of the Phenotypic Attribute and Trait (PATO)
ontology \cite{Gkoutos2005}. According to the PATO framework, a
phenotype can be decomposed, using an Entity-Quality model, into an
affected entity and a quality that characterizes {\em how} the entity
is affected \cite{Gkoutos2005}. Such decompositions have been created
for several widely used phenotype ontologies \cite{Mungall2010,
  Gkoutos2009b, obml2011h1}, and are being applied together with
methods for reusing knowledge contained in anatomy ontologies
\cite{Mungall2010, Hoehndorf2010phene}.

@@

% To illustrate these challenges, consider physiological processes of
% the heart. One of the heart's functions is {\em Heart beating}, i.e.,
% a capability that is realized through processes of the type {\em Heart
%   beating}. {\em Blood} is a participant of {\em Heart beating}
% processes.  An abnormal phenotype of an organism could be that the
% {\em rate of heart beating} is increased. The intended meaning of such
% a description is that the number of {\em Heart beating} processes in a
% given time interval is higher than normal. Another important attribute
% of heart physiology is the fluid flow rate through the heart. For an
% abnormal phenotype such as {\em increased rate of fluid flow through
%   the heart}, the intended meaning could either be that the amount of
% fluid that is moved through the heart within a single heart beating
% process is increased or that the amount of fluid that is moved through
% the heart within a period of time is increased.

% Based on these examples, we can make several observations about
% process attributes. First, for a process like {\em Heart beating}, we
% can distinguish between single occurrences and processes in which {\em
%   Heart beating} occurs multiple times. Only the latter kind of
% process may have a {\em rate of heart beating}, while {\em Heart
%   beating} processes do not have such an attribute. Second, we can
% distinguish between abnormal fluid flow rates in {\em Heart beating}
% processes and rate of fluid flow through the heart within a given
% duration. Both may have entirely different underlying causes and it is
% therefore important to distinguish between them. Finally, we may be
% able to infer some phenotypes from others, thereby limiting the number
% of phenotypes that must be experimentally observed. For example, when
% the fluid flow rate in single heart beating processes is increased and
% the {\em rate of heart beating} is increased within a process $P$,
% then the rate of fluid flow through the heart will be increased for
% $P$.

Here, we present the foundations for an ontology of process
phenotypes. We present a theory in which several kinds of process
attributes can be distinguished so that normal and abnormal physiology
of biological systems can be formally characterized. We apply this
theory to cellular processes and create the Cell Phenotype Ontology
(CPO). The CPO is linked to reference ontologies for qualities,
biological processes, functions and cell components, and its prime
application is the unification of phenotypes on the cellular level
across different species as well as for annotation of cellular
phenotypes in domains in which no such ontology exists.

@@

PATO allows for the description of affected entities by combining
various ontologies that describe the entities that have been affected,
such as the various anatomical ontologies, GO \cite {Ashburner2000short},
the Cell Type Ontology \cite {Bard2005}, SO \cite {Eilbeck2005} etc
with the various qualities it provides for defining how these entities
were affected.  PATO can be used for annotation either directly in a
so called post-composed (post-coordinated) manner or for providing
formal (logical) definitions (equivalence axioms) to ontologies
containing a set of precomposed (pre-coordinated) phenotype terms. For
instance, to describe the decrease in the length of the sexual cycle
of female animals, we can combine the PATO term \textit{decreased
  duration} ({\tt PATO:0000499}) with the Gene Ontology term
\textit{estrous cycle} ({\tt GO:0042698}), whilst if such a term
existed in a pre-composed ontology (for example the {\tt MP:0009007}
term from the Mammalian Phenotype) it could be used to provide an
equivalence statement between that class and the above PATO-based
description.

\subsection{Axioms for physiology phenotypes}
We implement our theory of physiology phenotypes using OWL, a formal
language based on description logics. Using OWL, we formulate axioms
that can be used by automated reasoners to infer additional
information. Automated reasoning and the axioms we provide are
intended to satisfy three aims. First, we use the axioms to infer
information from background knowledge. In particular, we aim to
automatically generate a class hierarchy of physiology phenotypes when
an ontology of physiological processes, such as the GO, is provided as
background knowledge. Our second aim is to provide interoperability
with phenotype ontologies of other domains, including species-specific
phenotype ontologies that have been formalized using the EQ method
\cite{Mungall2010, Gkoutos2005}. Finally, our third aim is to query
physiology phenotypes based on physiological processes that are
affected or based on the way in which they are affected.

Our three aims rely on the possibility of using automated reasoning
over a resulting ontology of physiology phenotypes.  However, OWL is
an expressive formal language, and automated reasoning in OWL has a
high computational complexity (reasoning in OWL 2.0 is
2NEXPTIME-complete). Consequently, due to the exponential increase in
worst-case time complexity for automated reasoning in OWL, we would
not be guaranteed to achieve our aims once we consider more than very
few phenotype classes. In particular, using an ontology of the size of
GO, with more than XX,XXX classes for processes, as a foundation for
constructing an ontology of physiology phenotypes would not allow us
to achieve our aims through automated reasoning.

The OWL EL profile is a subset of OWL that significantly decreases the
expressivity of OWL and the resulting time complexity of automated
reasoning \cite{owlprofiles}. Highly efficient automated reasoners
have been developed for OWL EL which are capable of processing very
large ontologies \cite{Kazakov2011}. In order to achieve our aims and
utilize automated reasoning for large ontologies, we limit ourselves
to the OWL EL profile. As a consequence of the restriction to OWL EL,
we cannot make use of negation ({\tt not}), disjunction ({\tt or}),
universal quantification ({\tt forall}) and several other types of
operations in our axioms \cite{owlprofiles}.

The lack of expressivity in OWL EL requires a formulation of axioms so
that the inferences we desire (i.e., the subclass relations resulting
in the ontology's taxonomy) are maintained without using features of
OWL that go beyond OWL EL expressivity. Consequently, we formulate
{\em abnormality} and {\em absence} similarly to current
formalizations of EQ-based phenotype ontologies, without the use of
negation, disjunction or universal quantification
\cite{Mungall2010}. A detailed description of the axioms we implement
is available as supplementary material (Supplement 1).
\end{methods}

@@

Amongst abnormal physiology, we distinguish between abnormalities in a
{\em single occurrence} of a cell component's functioning and an
abnormal {\em pattern of multiple occurrences} of a cell component's
functioning. For example, abnormalities in cell division resulting in
{\em Aneuploidy} refer to abnormalities of {\em Cell division}
processes, while an {\em Increased rate of cell division} refers to an
abnormality in the pattern of occurrence of multiple cell division
processes. In the CPO we follow the GO and represent abnormalities in
the pattern of occurrence of $X$ as abnormalities of {\em Regulation
  of $X$} processes.  In particular, an {\em Increased rate of cell
  division} is not an attribute of {\em Cell division} processes, but
rather arises from the collection of all {\em Cell division} processes
that occur within an organism at a given time. In the
GO, {\em Regulation of $X$} processes refer to those processes that
determine how often and the way in which one or more $X$ processes
occur. Therefore, we assign the attribute of {\em Increased rate of
  cell division} to {\em Regulation of cell division} processes.

@@

A further domain that will greatly benefit from the CPO is {\em
  systems microscopy}, which aims to understand complex and dynamic
cellular systems by combining automated fluorescence microscopy, cell
microarray platforms, quantitative image analysis and data mining
\cite{Lock2010}.  If we consider some of the studies, which have been
published in this field in the last few years \cite{Neumann2010,
  Schmitz2010short, Fuchs2010}, the need for CPO becomes evident.  In the
three studies, live-cell imaging assays and RNAi knockdown were used
to generate phenotypic profiles that quantify the cellular response to
a given siRNA thus allowing identification of hundreds of genes
involved in diverse biological functions including cell division,
migration and survival.  In each study, several phenotypes were
detected and described by the authors without the use of ontologies or
controlled vocabularies, making the integration between datasets
extremely difficult. For example, it is evident that cell division
phenotypes were observed in all three datasets and referred to by
terms such as {\em Mitotic delay/arrest}, {\em Prolonged mitotic
  exit}, {\em Methaphase delay} and {\em Methaphase cells}). Without a
controlled vocabulary of cellular phenotypes, the overlap between such
phenotype descriptions is unclear.

@@

\bibliographystyle{plain}

\bibliography{/home/leechuck/Documents/papers/bibtex/lc}
